---
title: "When are Philosophy Articles Cited?"
abstract: |
  It's natural to believe that philosophy citations are typically to long ago pieces. We're still talking about philosophers from millenia ago. More strikingly, we're still talking about papers from half a century ago not as historical papers, but as part of the contemporary debate. But a systematic look at the citation data shows that these cases are outliers. Most citations are to recently published works. Surprisingly, this is less true in recent years than it used to be. The effect of electronic publishing and communication has been to make citations, on average, older. After we adjust for the typical age of philosophy citations, and this changing trend, it turns out that the 2000s were a particularly influential time in philosophy publishing. Articles published in that decade are cited more than earlier or later articles, once we adjust for the typical times articles are cited, and the changing patterns of citation. This is arguably related to broad changes in the interests of philosophers, towards social philosophy, and epistemology.
execute:
  echo: false
  warning: false
date: today
bibliography: 
 - /Users/weath/Documents/quarto-articles/brian-quarto.bib
 - /Users/weath/Documents/citations-2025/cslbib.yaml
number-sections: true
keep-tex: true
author:
  - name: Anon
format:
  html:
    fig-format: png
    fig-width: 10
    fig-height: 7
  docx:
    reference-doc: my-template.docx
  pdf: 
    geometry: "left=1in,right=1in,top=1in,bottom=1in"
    mathfont: EB Garamond Math
    mainfont: EB Garamond Math
    sansfont: EB Garamond SemiBold
    mainfontoptions: 
      - ItalicFont=EB Garamond Italic
      - BoldFont=EB Garamond SemiBold
    fontsize: 12pt
    linkcolor: black
    urlcolor: black
    colorlinks: false
    linestretch: 1.75
    link-citations: true
    output-file: "Age of citations.pdf"
    indent: 1in
    include-in-header:
      text: |
        \setlength\heavyrulewidth{0ex}
        \setlength\lightrulewidth{0ex}
---

```{r}
#| label: loader
#| cache: false

require(tidyverse)
require(slider)
require(stringr)
require(knitr)
require(lsa)
require(wesanderson)

load("/Users/weath/Documents/citations-2025/philo_bib_through_2024.RData")
load("/Users/weath/Documents/citations-2025/philo_cite_through_2024.RData")
load("/Users/weath/Documents/articles/apc/active_journal_list.RData")

if(knitr::is_latex_output()) {
  knitr::opts_chunk$set(dev = 'cairo_pdf')
}

# Graph Themes
old <- theme_set(theme_minimal())
theme_set(old)
theme_update(plot.title = element_text(family = "Scala Pro", size = 24, face = "bold"),
             plot.subtitle = element_text(family = "Scala Sans Pro", size = 30),
             axis.text = element_text(family = "Scala Sans Pro", size = 18),
             title = element_text(family = "Scala Sans Pro", size = 18),
             plot.background = element_rect(fill = "#F9FFFF"),
             panel.background = element_rect(fill = "white"),
             legend.background = element_rect(fill = "#F9FFFF"),
             panel.grid = element_line(color = "grey92"),
             legend.text = element_text(family = "Scala Sans Pro", size = 20),
             strip.text = element_text(family = "Scala Sans Pro", size = 20),
             legend.key.spacing.y = unit(0.5, 'lines'),
             legend.key.spacing.x = unit(1, 'cm')
  )

if(knitr::is_latex_output()) {
theme_update(axis.title = element_text(family = "EB Garamond", size = 11),
             plot.title = element_text(family = "Europa-Bold", size = 14),
             plot.subtitle = element_text(family = "EB Garamond", size = 11),
             axis.text = element_text(family = "EB Garamond", size = 10),
             plot.background = element_rect(fill = "white"),
             panel.background = element_rect(fill = "white"),
             legend.background = element_rect(fill = "white"),
             panel.grid = element_line(color = "grey92"),
             legend.text = element_text(family = "EB Garamond", size = 11),
             strip.text = element_text(family = "EB Garamond", size = 12),
             legend.key.spacing.y = unit(-0.3, 'lines'),
             legend.key.spacing.x = unit(0, 'cm')
  )
}
```

```{r}
#| label: buildgraphs
#| cache: true

start_year <- 1955
end_year <- 2024
min_data <- 5

# New attempt
# Two categories: available and typical
# Available means published before citing article
# Typical means published 3-10 years before citing article
# The 3 is because weird things have happened with recent cites in recent years

typical_low <- 3
typical_high <- 10

# This sets the color for one-color graphs

point_col <- wes_palette("AsteroidCity1")[3]

active_philo_bib <- philo_bib_through_2024 |>
  filter(year >= start_year, year <= end_year)

active_philo_cite <- philo_cite_through_2024 |>
  filter(
    id %in% active_philo_bib$id,
    refs %in% active_philo_bib$id
  )

citation_tibble <- active_philo_cite |>
  as_tibble() |>
  rename(new = id, old = refs) |>
  left_join(active_philo_bib |>
            select(id, year), 
            by = c("old" = "id")) |>
  rename(old_year = year)  |>
  left_join(active_philo_bib |>
            select(id, year), by = c("new" = "id")) |>
  rename(new_year = year) |>
  filter(old_year >= start_year,
         new_year <= end_year,
         old_year >= start_year,
         new_year <= end_year)

# Find the highly cited articles, and count their citations separately

high_threshold <- 15

highly_cited <- citation_tibble |>
  group_by(old) |>
  tally(name = "citations") |>
  filter(citations >= high_threshold) |>
  rename(id = old)

highly_cited_per_year <- active_philo_bib |>
  filter(id %in% highly_cited$id) |>
  group_by(year) |>
  tally(name = "high_articles") 

# Now a tibble of how many times articles in year x are cited in year y

year_in_year_out <- citation_tibble |>
  filter(old_year >= 1956) |>
  group_by(old_year, new_year) |>
  tally(name = "citations") |> # Now add the 'missing' pairs
  ungroup() |>
  complete(old_year, new_year, fill = list(citations = 0)) |>
  left_join(citation_tibble |>
              group_by(old) |>
              filter(n() >= high_threshold) |>
              group_by(old_year, new_year) |>
              tally(name = "high_citations") |> # Now add the 'missing' pairs
              ungroup() |>
              complete(old_year, new_year, fill = list(high_citations = 0)),
            by = c("old_year", "new_year")) |>
  replace_na(list(high_citations = 0)) |>
  mutate(low_citations = citations - high_citations)

# This works out how many citations there are each year to 3-10 year old articles

citations_in_typical_year <- year_in_year_out |>
  mutate(age = new_year - old_year) |>
  filter(age >= typical_low, age <= typical_high) |>
  group_by(new_year) |>
  summarise(typical_citations = sum(citations)) 

# I'm going to count the 'typical' articles as those published between 3 and 10 years before the citing year
# The 'available' articles are those published before the time

# Tibble for number of publications each year, and cumulative, or 'available'

articles_per_year <- active_philo_bib |>
  rename(old_year = year) |>
  group_by(old_year) |>
  tally(name = "articles") |>
  mutate(available = cumsum(articles)) |>
  mutate(typical_articles = slide_dbl(articles, sum, .before  = typical_high) - slide_dbl(articles, sum, .before = typical_low - 1)) |>
  filter(old_year >= 1956) |>
  left_join(highly_cited_per_year, by = c("old_year" = "year")) |>
  mutate(low_articles = articles - high_articles)

articles_per_year_plot <- articles_per_year |>
  ggplot(aes(x = old_year, y = articles)) +
  geom_point(color = point_col) +
  xlab(element_blank()) +
  ylab("Number of indexed articles")

typical_plot <- articles_per_year |>
  ggplot(aes(x = old_year, y = typical_articles)) +
  geom_point(color = point_col) +
  xlab(element_blank()) +
  ylab("Number of typical indexed articles")

# Same for citations

all_citations_per_year <- citation_tibble |>
  group_by(new_year) |>
  tally(name = "citations") 

all_citations_per_year_plot <- all_citations_per_year |>
  ggplot(aes(x = new_year, y = citations)) +
  geom_point(color = point_col) +
  xlab(element_blank()) +
  ylab("Citations to indexed articles")

typical_citations_per_year <- citation_tibble |>
  filter(new_year >= old_year + typical_low, new_year <= old_year + typical_high) |>
  group_by(new_year) |>
  tally(name = "citations") 

typical_citations_per_year_plot <- typical_citations_per_year |>
  ggplot(aes(x = new_year, y = citations)) +
  geom_point(color = point_col) +
  xlab(element_blank()) +
  ylab("Citations to indexed articles from typical years")


# Outbound citations

outbound_citations <- left_join(
  articles_per_year,
  all_citations_per_year,
  by = c("old_year" = "new_year")
) |>
  mutate(outbound_rate = citations/articles) |>
  mutate(outbound = round(outbound_rate, 2))

outbound_citations_plot <- outbound_citations |>
  filter(old_year != 1955) |>
  ggplot(aes(x = old_year, y = outbound)) +
  geom_point(color = point_col) +
  xlab(element_blank()) +
  ylab("Outbound citations per indexed articles")

# Citations per typical article

available_citation_rate_per_year <- available_citations_per_year |>
  left_join(articles_per_year, by = c("new_year" = "old_year")) |>
  #filter(new_year >= start_year + typical_high) |>
  left_join(citations_in_available_year, by = "new_year") |>
  mutate(mean_cites = available_citations/available)

available_citation_rate_per_year_plot <- available_citation_rate_per_year |>
  ggplot(aes(x = new_year, y = mean_cites)) +
  geom_point(color = point_col) +
  xlab(element_blank()) +
  ylab("Annual citation rate of available articles.")

typical_citation_rate_per_year <- typical_citations_per_year |>
  left_join(articles_per_year, by = c("new_year" = "old_year")) |>
  #filter(new_year >= start_year + typical_high) |>
  left_join(citations_in_typical_year, by = "new_year") |>
  mutate(mean_cites = typical_citations/typical_articles)

typical_citation_rate_per_year_plot <- typical_citation_rate_per_year |>
  ggplot(aes(x = new_year, y = mean_cites)) +
  geom_point(color = point_col) +
  xlab(element_blank()) +
  ylab("Annual citation rate of typical articles.")

# All citations to typical articles in a year
ct_all <- citation_tibble |>
  filter(new_year >= old_year + typical_low, new_year <= old_year + typical_high) |>
  group_by(new_year) |>
  tally(name = "typical_citations")

age_effect_tibble <- year_in_year_out |>
  filter(old_year >= start_year, old_year <= end_year + 1 - min_data, new_year >= start_year + typical_high) |>
  filter(new_year >= old_year) |>
  left_join(
    select(
      articles_per_year, 
      old_year, 
      articles,
      high_articles,
      low_articles), 
    by = "old_year") |>
  left_join(
    select(
      articles_per_year, 
      old_year, 
      typical_articles), 
    by = c("new_year" = "old_year")) |>
  left_join(ct_all, by = "new_year") |> 
  mutate(age = new_year - old_year) |>
  mutate(cite_ratio = (citations/articles)/(typical_citations/typical_articles)) |>
  mutate(high_cite_ratio = (high_citations/high_articles)/(typical_citations/typical_articles))  |>
  mutate(low_cite_ratio = (low_citations/low_articles)/(typical_citations/typical_articles)) 

age_effect_tibble_plot <- age_effect_tibble |>
  filter(old_year >= start_year + 1, old_year <= end_year - min_data, new_year >= start_year) |>
  ggplot(aes(x = new_year, y = cite_ratio)) +
  geom_point(size = 0.25, color = point_col) +
  facet_wrap(~old_year, ncol = 6) +
  xlab(element_blank()) +
  ylab(element_blank()) +
  theme(axis.text = element_text(size = 10),
        strip.text = element_text(size = 12))

age_effect_grouped <- age_effect_tibble |>
  filter(new_year >= old_year) |>
  filter(new_year <= old_year + end_year - start_year + 1 - min_data) |>
  mutate(age = new_year - old_year) |>
  group_by(age) |>
  summarise(mean_effect = mean(cite_ratio),
            high_mean_effect = mean(high_cite_ratio),
            low_mean_effect = mean(low_cite_ratio))

age_effect_tibble_adj <- age_effect_tibble |>
  mutate(age = new_year - old_year) |>
  filter(age <= end_year - start_year - min_data) |>
  left_join(age_effect_grouped, by = "age")

age_effect_grouped_plot <- age_effect_grouped |>
  ggplot(aes(x = age, y = mean_effect)) +
  geom_point() +
  xlab("Article age") +
  ylab("Mean citation ratio")

year_by_year_with_effect <- year_in_year_out |>
  filter(new_year >= old_year) |>
  filter(new_year <= end_year) |>
  filter(old_year >= start_year, old_year <= end_year - min_data + 1, new_year >= start_year + typical_high) |>
  mutate(age = new_year - old_year) |>
  filter(age <= end_year - start_year - min_data) |>
  left_join(age_effect_grouped, by = "age") |>
  left_join(
    select(
      age_effect_tibble, old_year, new_year, cite_ratio
    ), by = c("old_year", "new_year")
  ) |>
  mutate(surplus = cite_ratio - mean_effect) |>
  arrange(-surplus)

# The next one calculates the difference between each year and the average. 
# But this has odd effects at the periphery, and compares each year to something it is part of.
# Below, in yiyo_extended, I try to work out what happens when each year is compared to the other years
# This is more work because you have to calculate the 'other years' value again each time

year_by_year_average <- year_by_year_with_effect |>
#  filter(age <= 7) |>
#  filter(old_year != 1973) |>
  group_by(old_year) |>
  summarise(mean_surplus = mean(surplus))

year_by_year_average_plot <- year_by_year_average |>
  mutate(rolling = slide_mean(mean_surplus, before = 4, after = 4)) |>
  ggplot(aes(x = old_year, y = mean_surplus)) +
  geom_line(
    aes(x = old_year, y = rolling),
    linewidth = 0.5,
    alpha = 0.5,
    color = point_col
  ) +
  geom_point(color = point_col)  +
  xlab(element_blank()) +
  scale_x_continuous(breaks = (98:100)*20) +
  ylab("Mean annual citations above average") +
  scale_y_continuous(labels = scales::percent)

year_by_year_average_plot_short <- year_by_year_with_effect |>
  filter(age <= 7) |>
#  filter(old_year != 1973) |>
  group_by(old_year) |>
  summarise(mean_surplus = mean(surplus)) |>
  mutate(rolling = slide_mean(mean_surplus, before = 4, after = 4)) |>
  ggplot(aes(x = old_year, y = mean_surplus)) +
  geom_point(color = point_col)  +
  geom_line(
    aes(x = old_year, y = rolling),
    linewidth = 0.5,
    alpha = 0.5,
    color = point_col
  ) +
  xlab(element_blank()) +
  scale_x_continuous(breaks = (98:100)*20) +
  ylab("Mean annual citations above average") +
  scale_y_continuous(labels = scales::percent)

year_by_year_average_plot_long<- year_by_year_with_effect |>
  filter(age > 7) |>
#  filter(old_year != 1973) |>
  group_by(old_year) |>
  summarise(mean_surplus = mean(surplus)) |>
  mutate(rolling = slide_mean(mean_surplus, before = 4, after = 4)) |>
  ggplot(aes(x = old_year, y = mean_surplus)) +
  geom_point(color = point_col)  +
  geom_line(
    aes(x = old_year, y = rolling),
    linewidth = 0.5,
    alpha = 0.5,
    color = point_col
  ) +
  xlab(element_blank()) +
  scale_x_continuous(breaks = (98:100)*20) +
  ylab("Mean annual citations above average") +
  scale_y_continuous(labels = scales::percent)

#print(year_by_year_average_plot)

effect_by_age_average <- function(early, late){
  age_effect_tibble |>
    filter(age >= early, age <= late) |>
    #    add_count(old_year, name = "data_points") |>
    #    filter(data_points >= min_data) |>
    group_by(old_year) |>
    summarise(mean_ratio = mean(cite_ratio)) |>
    ggplot(aes(x = old_year, y = mean_ratio)) +
    geom_point() +
    geom_smooth() +
    xlab(element_blank()) +
    ylab(element_blank()) +
    labs(title = case_when(
      early == late ~ paste0("Citation ratio at age ", early),
      TRUE ~ paste0("Mean citation ratio from ages ",early," to ",late)))
}

effect_by_age_facet <- function(early, late){age_effect_tibble |>
    filter(age>= early, age <= late) |>
    ggplot(aes(x = old_year, y = cite_ratio)) +
    geom_point() + geom_smooth() +
    facet_wrap(~age, ncol = 4)
}

year_to_mean_plot <- function(the_year){
  age_effect_tibble_adj |>
    filter(old_year == the_year) |>
    ggplot(aes(x = age, y = cite_ratio)) +
    geom_point(size = 2, alpha = 1, color = hcl(h = (the_year-1975)*(360/43)+15, l = 65, c = 100)) +
    # geom_jitter(aes(size=(old_year==2008 | old_year == 1985), shape = (old_year==2008)), alpha = 1) +
    #  geom_jitter(aes(size=(old_year %in% c(1978, 1980, 1985, 1987)), alpha = 1)) +
    # scale_size_manual(values=c(0.3,2)) +
    xlab("Age of cited articles") +
    ylab("Citation ratio") +
    geom_line(aes(x = age, y = mean_effect), color = point_col) +
    geom_point(aes(x = age, y = mean_effect), color = point_col, size = 0.4) +
    theme(legend.position = "none")
}

```

```{r}
#| label: calculate-variables

citations_1956 <- scales::label_comma()(
  filter(all_citations_per_year, new_year == 1956)$citations)
citations_2024 <- scales::label_comma()(
  filter(all_citations_per_year, new_year == 2024)$citations)

number_of_articles <- scales::label_comma()(nrow(active_philo_bib))
number_of_citations <- scales::label_comma()(nrow(active_philo_cite))

synthese_2021 <- scales::label_comma()(
  nrow(
    filter(
      active_philo_bib,
      year == 2021,
      journal == "Synthese"
    )
  )
)
synthese_2022 <- scales::label_comma()(
  nrow(
    filter(
      active_philo_bib,
      year == 2022,
      journal == "Synthese"
    )
  )
)
```

# Introduction {#sec-introduction}

This paper is about the patterns of citations of philosophy journal articles in philosophy journals. Obviously philosophy journals cite more things than philosophy journals, and just as obviously philosophy journal articles get cited in other places. But looking just at journal-to-journal citations allows us to get a citation set that is relatively complete, and hence make some systematic generalisations about the way articles are cited over time. It turns out some of these generalisations are surprising.

Before looking at the data, here are two things I believed about philosophy citations. First, philosophers tend to cite very old papers. We still regularly teach a number of papers over half a century old in introductory classes; e.g., @WOSA1969Y444700002, @WOSA1971Y116900003, @WOSA1972Z066400001, and @10.2307_2025310. These aren't taught as history papers, but as early entries into the contemporary philosophical debate. While most papers aren't cited as much as these papers are, I thought the pattern that old papers keep being cited extended to their less famous counterparts. Second, the technological changes of the last quarter century meant that this practice was being slowly reversed. A series of technological innovations made it easier to cite newer and newer works. These innovations included the spread of email, the rise of preprint archives (e.g., arXiv, SSRN, PhilPapers), and eventually official preprints in things like EarlyView. So, I thought, citations should be getting younger, because the delay between publishing and getting widely known was removed.

Both of these thoughts were wrong.

On the first point, the generalisation I made from those famous papers was just wrong. Normal papers differ from famous papers not just in how often they are cited, but in the shape of their citations. The main evidence I'll use for this is something I'll call the _citation ratio_. The citation ratio of year _o_ in year _n_ is the mean number of citations, in year _n_, of articles published in year _o_, divided by the mean number of citations, in year _n_, of articles published in years _n_-10 to _n_-3. (I'll say much more about why I'm using this measure in what follows.) @fig-master-citation-ratio shows the average citation ratio for different _ages_, of citations, i.e., the number of years between _o_ and _n_.^[The graph also includes some 'jitter' to make the different points more easily visible. I've put each decade of original publication in a different colour; I'll break those out in @fig-decades-cite-ratio. The graph starts in 1975 because the data is much noisier before then, for reasons we'll get to below.]

```{r}
#| label: fig-master-citation-ratio
#| fig-cap: "Age effects from 1975 onwards on a single graph, with the overall average shown."

age_effect_post_1975 <- year_in_year_out |>
  filter(old_year >= 1975,
         new_year >= old_year) |>
  left_join(
    select(
      articles_per_year, 
      old_year, 
      articles,
      high_articles,
      low_articles), 
    by = "old_year") |>
  left_join(
    select(
      articles_per_year, 
      old_year, 
      typical_articles), 
    by = c("new_year" = "old_year")) |>
  left_join(ct_all, by = "new_year") |> 
  mutate(
    age = new_year - old_year,
    cite_ratio = (citations/articles)/(typical_citations/typical_articles),
    high_cite_ratio = (high_citations/high_articles)/
      (typical_citations/typical_articles),
    low_cite_ratio = (low_citations/low_articles)/
      (typical_citations/typical_articles)) |>
  ungroup() |>
  group_by(new_year - old_year) |>
  mutate(
    mean_cite_ratio = mean(cite_ratio),
    mean_high_cite_ratio = mean(high_cite_ratio),
    mean_low_cite_ratio = mean(low_cite_ratio)
  ) |>
  ungroup() |>
  mutate(decade = paste0(
    floor((old_year-5)/10),
    "5-",
    floor((old_year-5)/10)+1,
    "4"
  ))

ggplot(age_effect_post_1975, aes(x = new_year - old_year, 
                                 y = cite_ratio,
                                 color = decade)) +
  geom_jitter(size = 0.5, alpha = 0.7) +
  xlab("Age of cited articles") +
  ylab("Citation ratio") +
  geom_line(aes(x = age, y = mean_cite_ratio), color = point_col) +
  labs(color = element_blank()) +
  theme(legend.position = "none")
```

Each dot on that graph is a citation ratio for a particular pair of years; the line shows the average citation ratio for all pairs with the same age. The shape is unmistakable; articles get cited much much more when they are relatively young than when they are older.

The 'evidence' I gave for the opposite view in the introductory paragraph wasn't entirely wrong. If we redo @fig-master-citation-ratio just looking at articles which have 15 or more citations in philosophy journals. (This turns out to be a fairly small percentage of the sample.)

```{r}
#| label: fig-ageeffecteverything-high
#| fig-cap: "A version of @fig-master-citation-ratio just looking at highly cited articles"

ggplot(age_effect_post_1975, aes(x = new_year - old_year, 
                                 y = high_cite_ratio,
                                 color = decade)) +
  geom_jitter(size = 0.5, alpha = 0.7) +
  xlab("Age of cited articles") +
  ylab("Citation ratio") +
  geom_line(aes(x = age, y = mean_high_cite_ratio), color = point_col) +
  labs(color = element_blank()) +
  theme(legend.position = "none")
```

The numbers on the y-axis in @fig-ageeffecteverything-high are higher than in @fig-master-citation-ratio. That's not surprising; it just means highly cited articles get cited more frequently. What is striking is the different shape of the graphs. Typical philosophy articles, if they get cited at all, get cited soon after publication and they fade into obscurity. Highly cited articles keep getting cited decades after their publication. 

These results aren't a priori obvious; things could have turned out otherwise. It could have been that there were a trove of articles which were ignored after publication and then accrued five to ten citations a couple of decades later. There are some articles that were very frequently cited soon after publication but which are now largely ignored. (This happens most frequently in philosophy of science and in philosophy of mind, I think for different reasons in the two cases.) But these cases are outliers. Most of the articles that were influential soon after publication stay that way.

For the second point, we can simply break up @fig-master-citation-ratio by ten year chunks. In @fig-decades-cite-ratio I've taken the points by from @fig-master-citation-ratio, and grouped them into 'decades'. Because I'm working here with 1975-2024 data, the decades are 1975-1984, 1985-1994 etc. To make it easier to compare decades, I've removed the last one, where there isn't enough data, and removed all points with an age over 20.

```{r}
#| label: fig-decades-cite-ratio
#| fig-cap: "Citation ratio for different decades"
#| fig-subcap:
#|    - "1975-1984"
#|    - "1985-1994"
#|    - "1995-2004"
#|    - "2005-2014"
#| layout-ncol: 2

ten_year_graph <- function(x){
  temp <- age_effect_post_1975 |>
    filter(old_year >= x, old_year <= x + 9, 
           new_year >= old_year, new_year <= old_year + 20) |>
    group_by(new_year - old_year) |>
    mutate(mean_effect = mean(cite_ratio))
  
  temp |>
    ggplot(aes(x = age, y = cite_ratio, color = as.factor(old_year))) +
    geom_jitter(size = 0.5, alpha = 0.7) +
    xlab("Age of cited articles") +
    ylab("Citation ratio") +
    geom_line(aes(x = age, y = mean_effect), color = point_col) +
    geom_point(aes(x = age, y = mean_effect), color = point_col, size = 0.4) +
    theme(legend.position = "none") +
    ylim(0, 1.6)
}

ten_year_graph(1975)
ten_year_graph(1985)
ten_year_graph(1995)
ten_year_graph(2005)
```

There are three general trends across these graphs, especially after the second graph.

1. The peaks are getting later. In the first two graphs, the line is clearly heading down by age 5; in the last one it is barely off the peak.
2. The peaks are getting lower. In the last graph we barely see it cross 1.
3. The declines are much, much flatter. If you look around age 15 in the four graphs, you see the values rise steadily over time.

What all this means is that citations are getting older. While it's still true that articles from a year are (collectively) cited a more often from ages 2-5 than from ages 12-15, the difference between those two rates has fallen remarkably. The effect of technology on citations has been the complete opposite of what I expected.

The rest of this paper has two aims. 

First, I'm going to set out the methodology behind these graphs, go over the choices I made in building them, and argue that these were at least defensible choices. The intended conclusion is that these graphs really show what I say they do, that traditionally citations were mostly to very recent articles, but they are now more frequently to older articles. 

Second, I'm going to look citations from various years, after adjusting for these typical citation rates, and see which years have been more influential in the later literature. I suspect readers will not be surprised that the early 1970s stand out as being particularly influential. What might be more surprising is that the next most influential period, in terms of how often articles from then are cited compared to the overall trends, is the 2000s. There are a few possible reasons for this, but I suspect the main one is the rising importance at that time of epistemology. (This is something Eugenio @Petrovich2024 also found using a somewhat different data set.) More generally, looking at citations from different periods, and especially looking at which articles make up those citations, is a useful guide to the history of those periods. Most work on the history of analytic philosophy doesn't get beyond the early 1970s; this is an early attempt to quantify what happens in the years after the changes brought about by Kripke, Lewis, Rawls and others in those years.